---
title: "OBH paper draft"
author: "Elements Wilderness Program Research Team"
date: "9/19/2021"
output:
  pdf_document: default
  html_document: default
---


## Data Setup  

```{r setup, include = FALSE}
# Removes all code from the output

library(tidyverse)
library(janitor) # For tabyl
library(knitr) # for kable

knitr::opts_chunk$set(echo = FALSE)

getwd()

# Load OBH NATSAP dataset and name "obh"
obh <- readRDS("C:/Users/cullo/OneDrive/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/rds/obh_for_logistic.rds")


```

**To view the cleaning script for this file, please visit "OBH cleaning.R" on Github**


# Frequency tables  

<br>

## Y/N response variables  

```{r frequency tables}


# Build a table of useful variables
# Includes: gender, all boolean(5:19), Transport(24), age, last_record, and likert bins


freq_tabs_list <- lapply(obh[,c(1, 5:20, 25)], tabyl) 

# Bring this list of tables together into a central table
freq_tab1 <- data.table::rbindlist(freq_tabs_list, fill = TRUE, idcol=TRUE)

# Tidy up
# library(tidyverse)
freq_tab <- freq_tab1 %>% rename(variable = `.id`, response = `X[[i]]`) %>% 
  mutate(percent = round(percent*100, 2)) %>% 
  select(-valid_percent)


kable(freq_tab)

# names(obh)
# mean(obh$age_intake)
# sd(obh$age_intake)

```


Now remove any variables that had less than 5% "yes" responses

```{r freq_select vars}
# Remove variables that have less than 5% response rate
obh <- obh %>% select(-c(bi_polar, eating_disorder, gender_dysphoria, learning_disability, ocd_disorders, personality_disorders, psychotic_disorders))

# New freq table with updated inputs
freq_tabs_list1 <- lapply(obh[,c(1, 5:13, 18)], tabyl) 

# Bring this list of tables together into a central table
freq_tab_rm <- data.table::rbindlist(freq_tabs_list1, fill = TRUE, idcol=TRUE)

# Tidy up
freq_tab_rm <- freq_tab_rm %>% rename(variable = `.id`, response = `X[[i]]`) %>% 
  mutate(percent = round(percent*100, 2)) %>% 
  select(-valid_percent)

kable(freq_tab_rm)

```





## Intake Diff Hist

```{r z_dif setup}

# from obh$rel01_dif_z

obh <- obh %>% mutate(z_dif_cat = 
                        case_when(rel01_dif_z <= -1 ~ -1,  # student high
                                  rel01_dif_z > -1 & rel01_dif_z < 1 ~ 0, # same
                                  rel01_dif_z >= 1 ~ 1))   # parent high

# Code this variable non-numerically as well
obh <- obh %>% mutate(z_dif_levels = 
                        case_when(rel01_dif_z <= -1 ~ "stuH",  # student high
                                  rel01_dif_z > -1 & rel01_dif_z < 1 ~ "same", # same
                                  rel01_dif_z >= 1 ~ "parH"))   # parent high

obh$z_dif_levels <- relevel(as.factor(obh$z_dif_levels), "same")


```


```{r Introduce Intake Difference}
## Define INTAKE DIFFERENCE----
# Convert client and parent intake scores to z-scores and find the difference
# to understand if disagreement of severity of problem impacts program success

# Convert to Z scores
obh <- obh %>% 
  # Client Intake:: (client_A0_z)
  mutate(client_A0_z = (client_YOQ_20SR_A0_SCORE - mean(client_YOQ_20SR_A0_SCORE, na.rm = TRUE))/ 
           sd(client_YOQ_20SR_A0_SCORE, na.rm=TRUE)) %>% 
  # Parent Intake (rel01_A0_z)
  mutate(rel01_A0_z = (rel01_YOQ_201_A0_SCORE - mean(rel01_YOQ_201_A0_SCORE, na.rm = TRUE))/ 
           sd(rel01_YOQ_201_A0_SCORE, na.rm=TRUE)) 
# Calculate difference in z scores
## RELATIVE01 - CLIENT and REL01 - CLIENT
obh <- obh %>% 
  mutate(rel01_dif_z = (rel01_A0_z - client_A0_z))

# Bin intake difference z scores based on standard deviations from 0
obh <- obh %>% mutate(z_dif_cat = 
                        case_when(rel01_dif_z <= -1 ~ -1,  # student high
                                  rel01_dif_z > -1 & rel01_dif_z < 1 ~ 0, # same
                                  rel01_dif_z >= 1 ~ 1), # parent high
                      z_dif_levels =
                        case_when(rel01_dif_z <= -1 ~ "stuH",  # student high
                                  rel01_dif_z > -1 & rel01_dif_z < 1 ~ "same", # same
                                  rel01_dif_z >= 1 ~ "parH"))   # parent high
# Set reference level to "same" for regression
obh$z_dif_levels <- relevel(as.factor(obh$z_dif_levels), "same")
```






```{r}
# goal to overlap client_A0_z rel01_A0_z then also rel01_dif_z?

intake_dif_plot <- ggplot(obh) +
  geom_density(aes(x = client_YOQ_20SR_A0_SCORE), linetype = "dashed", size = 1) +
  geom_density(aes(x = rel01_YOQ_201_A0_SCORE), size = 1) +
  theme_minimal() +
  xlab("YOQ Intake Score")+
  labs(title = "Intake distribution difference", caption = "dashed = client")


z_intake_dif_plot <- ggplot(obh) +
  geom_density(aes(x = client_A0_z), linetype = "dashed", size = 1) +
  geom_density(aes(x = rel01_A0_z), size = 1) +
  theme_minimal() +
  xlab("YOQ Intake Z Score")+
  labs(title = "Z Score Intake distribution difference", caption = "dashed = client") +
  geom_vline(xintercept = 1)+
  geom_vline(xintercept = -1)


# ggsave("C:/Users/applesaint/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/fig/intake difference plot.jpg", intake_dif_plot, width = 10, height = 7)
# ggsave("C:/Users/applesaint/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/fig/z score intake difference plot.jpg", z_intake_dif_plot, width = 10, height = 7)
# 


only_z_dif_plot <- ggplot(obh) +
  geom_density(aes(x = rel01_dif_z), size = 1) +
  theme_minimal() +
  xlab("Parent - Student Z Score")+
  labs(title = "Z Score Intake distribution difference", caption = "dashed = client") +
  geom_vline(xintercept = 1, linetype = "dashed")+
  geom_vline(xintercept = -1, linetype = "dashed")


# ggsave("C:/Users/applesaint/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/fig/only z difference plot.jpg", only_z_dif_plot, width = 10, height = 7)

obh %>% group_by(z_dif_cat) %>% tally()
# 379 missing

colSums(is.na(obh))

```




<br>



## Change Category formula

```{r change cat formula, echo = TRUE}
# Note, this is already done preemptively on rds file, just to see process

example_formula <- obh %>% 
  mutate(change_cat_client =
           case_when(delta_client <= -18 & client_YOQ_20SR_D0_SCORE <= 46 ~ "rec", # recovered
                     delta_client <= -18 & client_YOQ_20SR_D0_SCORE >= 47 ~ "imp", # improved
                     delta_client >= -17 & delta_client <= 0 & 
                       client_YOQ_20SR_D0_SCORE >= 47 ~ "unch", # unchanged
                     delta_client >= 1 & client_YOQ_20SR_D0_SCORE >= 47 ~ "det", # deteriorated
                     delta_client >= -17 & delta_client <= 0 & client_YOQ_20SR_D0_SCORE <= 46 ~ "nocl", # non-clinical
                     delta_client >= 1 & client_YOQ_20SR_D0_SCORE <= 46 ~ "nocl")) 


```


## Change category frequency  
```{r change_cat_tab}


# Set up frequency table of clinical change categories
change_cat_tab <- obh %>% group_by(change_cat_client) %>% tally() %>% 
  mutate(change_cat_client = recode(change_cat_client, rec = "recovered", imp = "improved",
                                    unch = "unchanged", nocl = "non_clinical", det = "deteriorated")) %>% 
  mutate(perc = round(n/2470*100,2))


kable(change_cat_tab)

```
For the purpose of this study, we are going to dummy code an RCI variable



## RCI  
```{r RCI}
## Run again for RCI


# obh$rci <- ifelse(obh$delta_client <= - 18, 1, 0)

rci_tab <- obh %>% group_by(rci) %>% tally() %>% mutate(perc = round(n/2470*100,2))

kable(rci_tab)

# And create ric_mo6
obh$rci_mo6 <- ifelse(obh$delta_client_mo6 <= - 18, 1, 0)



```




## Means and multi-response tables:  
age, clinical change category, program_need, change_desire, prior_prog, progress_need, intake_difference
```{r means and multi_response tabs}

tab1 <- data.frame("var" = c("age_intake", "program_need", 
                             "change_desire", "prior_prog", "progress_need"),
                   "mean" = c(mean(obh$age_intake), 
                              mean(obh$program_need, na.rm = TRUE), 
                              mean(obh$change_desire, na.rm = TRUE),
                              mean(obh$prior_prog, na.rm = TRUE), 
                              mean(obh$progress_need, na.rm = TRUE)),
                   "sd" = c(sd(obh$age_intake), 
                              sd(obh$program_need, na.rm = TRUE), 
                              sd(obh$change_desire, na.rm = TRUE),
                              sd(obh$prior_prog, na.rm = TRUE), 
                              sd(obh$progress_need, na.rm = TRUE)),
                   "n" = c(sum(!is.na(obh$age_intake)),
                           sum(!is.na(obh$program_need)),
                           sum(!is.na(obh$change_desire)),
                           sum(!is.na(obh$prior_prog)),
                           sum(!is.na(obh$progress_need))
                           ))



kable(tab1)


```

## Descriptive Stats on YOQ Scores  

```{r yoq means}
yoq_mean <- data.frame("var" = c("client_intake", "rel01_intake", "client_discharge", "client_change"),
                   "mean" = c(mean(obh$client_YOQ_20SR_A0_SCORE), 
                              mean(obh$rel01_YOQ_201_A0_SCORE, na.rm = TRUE), 
                              mean(obh$client_YOQ_20SR_D0_SCORE, na.rm = TRUE),
                              mean(obh$delta_client, na.rm = TRUE)),
                   "sd" = c(sd(obh$client_YOQ_20SR_A0_SCORE), 
                              sd(obh$rel01_YOQ_201_A0_SCORE, na.rm = TRUE), 
                              sd(obh$client_YOQ_20SR_D0_SCORE, na.rm = TRUE),
                              sd(obh$delta_client, na.rm = TRUE)),
                   "n" = c(sum(!is.na(obh$client_YOQ_20SR_A0_SCORE)),
                           sum(!is.na(obh$rel01_YOQ_201_A0_SCORE)),
                           sum(!is.na(obh$client_YOQ_20SR_D0_SCORE)),
                           sum(!is.na(obh$delta_client))
                           ))

kable(yoq_mean)
```


Variables under consideration:  

* gender  

* adopted  

* attention_issue  

* autism  

* conduct_disorder  

* depression  

* substance_abuse  

* trauma_related_issue  

* program_need  

* change_desire  

* prior_prog  

* progress_need  

* Routine Monitoring Status

* transport  

* Rel01 intake difference z score  




# Chi-Squared on groups 

Do we want to run this with our dummy variable or full clinical change categories?   
Think we'll want to run rci because that will be what we are feeding into the logistic regression

```{r chisq}
# dummy code intake difference score
# obh$change_cat_1sd <- ifelse(obh$rel01_dif_cat == 3, 1, 0)




# Subset just tested variables
obh_sel <- obh %>% select(rci, gender, adopted, attention_issue, autism, conduct_disorder, depression, substance_abuse, trauma_related_issue, rom, program_need, change_desire, prior_prog, progress_need, transport, z_dif_cat)



combos <- combn(ncol(obh_sel),2)
# Only want 1:13 ie. combos with rci
combos <- combos[,1:15]
x2_test <- plyr::adply(combos, 2, function(x) {
  test <- chisq.test(obh_sel[, x[1]], obh_sel[, x[2]])

  out <- data.frame("Row" = colnames(obh_sel)[x[1]]
                    , "Column" = colnames(obh_sel[x[2]])
                    , "Chi.Square" = round(test$statistic,3)
                    ,  "df"= test$parameter
                    ,  "p.value" = round(test$p.value, 3)
                    )
  return(out)

})  



kable(x2_test)

```


Separate chi-squared tests showed a number of variables having significant impact on the program having a positive impact on a student or not: gender, attention_issue, autism, depression, program_need, progress_need and transport.  *Will include all these numbers in paper, but here are significant variables:*  

```{r chi freq tab}

obh_sel1 <- obh_sel %>% select(-c(program_need, change_desire, prior_prog, progress_need))
obh_long <- gather(obh_sel1, key = "variable", value = "response", c(2:12))


table(obh$rci)
# 0 = 1122, 1 = 1348


# Put all into one table
x2_des <- obh_long %>% 
  group_by(variable, response, rci) %>% tally() %>%
  pivot_wider(names_from = rci, values_from = n) %>% 
  na.omit() %>% 
  rename(no_rci = `0`, rci = `1`)

# Add total
x2_des$total <- x2_des$no_rci + x2_des$rci

# Add precent meet rci
x2_des$perc_met_rci <- round(x2_des$rci/x2_des$total *100, 2)



kable(x2_des)


## Merge chi-squared test (x2_test) to this

x2_test <- x2_test %>% select(-c(Row, X1))

chi_sq_merge <- merge(x2_des, x2_test, by.x = "variable", by.y = "Column")




## Save both these as csv files
# Not sure why having to use absolute path here, keeps changing to /script otherwise...
# write.csv(x2_des, "C:/Users/applesaint/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/output/x2_description_table.csv")
# write.csv(x2_test, "C:/Users/applesaint/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/output/x2 test table.csv")
# write.csv(chi_sq_merge, "C:/Users/applesaint/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/output/chi_sq results.csv")




```




# Mean of likert variables

```{r}
# Means of 0/1 for likert variables

likert_mean_tab <- obh %>% group_by(rci) %>% 
  summarise(program_need_mean = mean(program_need, na.rm = TRUE),
            program_need_sd = sd(program_need, na.rm = TRUE),
            change_desire_mean = mean(change_desire, na.rm = TRUE),
            change_desire_sd = sd(change_desire, na.rm = TRUE),
            prior_prog_mean = mean(prior_prog, na.rm = TRUE),
            prior_prog_sd = sd(prior_prog, na.rm = TRUE),
            progress_need_mean = mean(progress_need, na.rm = TRUE),
            progress_need_sd = sd(progress_need, na.rm = TRUE))


likert_mean_tab



```


```{r z difference table}

z_dif_tab <- obh %>% group_by(z_dif_cat, rci) %>% tally()

kable(z_dif_tab)

# Or just give mean difference z score based on rci

t1 <- obh %>% group_by(rci) %>% summarise(mean = mean(rel01_dif_z, na.rm=TRUE),
                                          sd = sd(rel01_dif_z, na.rm=TRUE))




```




# D0 Predictor Regression

```{r model_D0 logistic regression}

# Run model again with those select variables
obh_d0 <- obh %>% 
  select(gender, age_intake, adopted:transport, z_dif_levels, rci) %>% 
  filter(age_intake >= 13) %>% 
  na.omit()

# Removing na's brings data down to 2192 from 2418
model_D0 <- glm(rci ~ ., data = obh_d0, family = "binomial")
summary(model_D0)

model_D0_df <- as.data.frame(summary(model_D0)$coef)

model_D0_df$oddrat <- exp(coef(model_D0))


#(646 observations deleted due to missingness)

## Check for multicollinearity in model

# Correlation in all variables in our dataset

regclass::VIF(model_D0)


# write.csv(model_D0_df, file="C:/Users/applesaint/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/output/model_D0 Regression results.csv")

```



```{r D0 psuedo r^2}
## Calculate McFadden's psuedo R^2

## Now calculate the overall "Pseudo R-squared" and its p-value
model_D0_null <- model_D0$null.deviance/-2
model_D0_proposed <- model_D0$deviance/-2

## McFadden's Pseudo R^2 = [ LL(Null) - LL(Proposed) ] / LL(Null)
(model_D0_null - model_D0_proposed) / model_D0_null  
# .10, not great?

## The p-value for the R^2
1 - pchisq(2*(model_D0_proposed - model_D0_null), df=(length(model_D0$coefficients)-1))
# p-val very small
```


```{r D0 pass/fail regression}
## now we can plot the data
pred_d0 <- data.frame(rci_prob=model_D0$fitted.values, # sets the fitted vals to new column
                            rci=obh_d0$rci) # takes recovered value from original df and adds to this df
pred_d0 <- pred_d0[order(pred_d0$rci_prob, decreasing=FALSE),] # Probabilty order from low to high

pred_d0$rank <- 1:nrow(pred_d0) # calls row number to new column = "rank"

ggplot(data=pred_d0, aes(x=rank, y=rci_prob)) +
  geom_point(aes(color=rci), alpha=1, shape=4, stroke=2) +
  xlab("Index") +
  ylab("Predicted probability of Recovery")


pred_d0$pred_rci <- ifelse(pred_d0$rci_prob > .6, "1_pred", "0_pred") 
table(pred_d0$pred_rci, pred_d0$rci)




```



# D180 Predictor Regression


```{r model_D180 regression}
# Set up d180 dummy var if client has data at 6 months also

obh$d180 <- ifelse(!is.na(obh$client_YOQ_20SR_D180_SCORE), 1, 0)

# 
# Set up rci_d180 to give a rci for the discharge to 180 score
obh$rci_d180 <- ifelse(obh$delta_client_mo6 <= -18, 1, 0)

# Check to make sure all is working correctly
# t2 <- obh %>% select(d180, client_YOQ_20SR_D180_SCORE, client_YOQ_20SR_A0_SCORE, delta_client_mo6, rci_d180, rci)

# Run model again with those select variables
obh_d180 <- obh %>% 
  filter(d180 == 1) %>% 
  select(gender, age_intake, adopted:transport, z_dif_levels, rci_d180) %>% 
  filter(age_intake >= 13) %>% 
  na.omit()

# Removing na's brings data down to 2192 from 2418
model_D180 <- glm(rci_d180 ~ ., data = obh_d180, family = "binomial")
summary(model_D180)




## Check for multicollinearity in model

# Correlation in all variables in our dataset

regclass::VIF(model_D180)

# write.csv(as.data.frame(summary(model_D180)$coef), file="C:/Users/applesaint/Desktop/One Cloud/ELEMENTS/OBH_CF_CC/output/model_D180 Regression results.csv")



```



```{r D180 psuedo r^2}

## Calculate McFadden's psuedo R^2

## Now calculate the overall "Pseudo R-squared" and its p-value
model_D180_null <- model_D180$null.deviance/-2
model_D180_proposed <- model_D180$deviance/-2

## McFadden's Pseudo R^2 = [ LL(Null) - LL(Proposed) ] / LL(Null)
(model_D180_null - model_D180_proposed) / model_D180_null  
# .17, not great?

## The p-value for the R^2
1 - pchisq(2*(model_D180_proposed - model_D180_null), df=(length(model_D180$coefficients)-1))
# p-val very small





```
```{r D180 pass/fail regression}
## now we can plot the data
pred_d180 <- data.frame(rci_prob=model_D180$fitted.values, # sets the fitted vals to new column
                            rci=obh_d180$rci) # takes recovered value from original df and adds to this df
pred_d180 <- pred_d0[order(pred_d180$rci_prob, decreasing=FALSE),] # Probabilty order from low to high

pred_d180$rank <- 1:nrow(pred_d180) # calls row number to new column = "rank"

ggplot(data=pred_d180, aes(x=rank, y=rci_prob)) +
  geom_point(aes(color=rci), alpha=1, shape=4, stroke=2) +
  xlab("Index") +
  ylab("Predicted probability of Recovery")


pred_d180$pred_rci <- ifelse(pred_d180$rci_prob > .6, "1_pred", "0_pred") 
table(pred_d180$pred_rci, pred_d180$rci)




```





# t-tests
```{r t-tests}
# differences in discharge vs 6 months score
ttest1 <- t.test(df5$client_YOQ_20SR_D0_SCORE, df5$client_YOQ_20SR_D180_SCORE, paired = TRUE)

# Differnece in parent01 and studnet scores at intake
ttest2 <- t.test(df5$client_YOQ_20SR_A0_SCORE, df5$rel01_YOQ_201_A0_SCORE, paired = TRUE)


## differences in delta scores
ttest3 <- t.test(df5$delta_client, df5$delta_client_mo6, paired = TRUE)
# there is no significant difference between the change scores at discharge and the change scores at 6 months

```

1. No significant in YOQ scores when measured at discharge and at 6 months post discharge:  
```ttest1```  
2. There is a significant difference in YOQ scores of adolescents and parents at intake:
```ttest2```

3.  There is no significant difference between the change scores at discharge and the change scores at 6 months (also ran logistic on improved (55, 13%) vs deteriorated (34, 8%) at 6 months and saw no significant predictors, but don't know if we want to report that)
```ttest3``` 

# Hypothesis 3: DOA-DOD vs DOA-MO6
```{r}
table(obh$rci_mo6, obh$rci)

obh %>% group_by(rci, rci_mo6) %>% tally()


# Run chi-squared?

chisq.test(obh$rci, obh$rci_mo6)

# Shows that this is a statistical difference between those with RCI at DOA and those with RCI at MO6
# Thus we cannot deuce the same things are happening with those folks

# Or make another variables same_at_mo6
obh$same_rci_mo6 <- ifelse(obh$rci == obh$rci_mo6, 1, 0)
table(obh$same_rci_mo6)


```


# Test Stu_high 

```{r}
obh$z_dif_levels

# Want to compare all in stuH (or, rather, only those that are about 1 sd from mean of total scores)

mean(obh$client_YOQ_20SR_A0_SCORE) # 71.09
sd(obh$client_YOQ_20SR_A0_SCORE)   # 33.43

# So to meet the cutoff of what we are trying to study, must only look at clients with scores at or above 104

obh$stu_high_intake <- ifelse(obh$client_YOQ_20SR_A0_SCORE > 104, 1, 0)

table(obh$stu_high_intake, obh$z_dif_levels)
# In these different groups, 145 of "same" are high at intake, and 204 of "stuH" are high at intake

stu_high <- obh %>% filter(stu_high_intake == 1, z_dif_levels != "parH")


tab_stu <- table(stu_high$rci, stu_high$z_dif_levels)

chisq.test(stu_high$rci, stu_high$z_dif_levels)

# Chi-squared says there is somthing going on between this group more than just random chance
```







